MPA Gap Analysis: IUCN and AquaMaps

### 2016-02-25 mpa_analysis/mpa_analysis2.Rmd

library(rgdal)
library(maptools)
library(raster)
library(readr)
library(tidyr)
library(dplyr)
library(stringr)

@@

  rast_empty <- raster(ext = extent(rast_loiczid), resolution = 0.01)
  rast_base  <- raster::area(rast_empty) ### takes a while...

@@

### as long as we're here, let's set up some objects for WGS 84 projection
p4s_wgs84 <- '+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'
crs_wgs84 <- CRS(p4s_wgs84)

rast_marine_file <- file.path(dir_git, 'mpa_analysis/rast_marine.tif')
reload <- FALSE
if(reload == TRUE) {
### import OHI regions polygons, to determine spatially-marine areas
  poly_marine_file <- file.path(dir_anx, 'explore_am_v_iucn/mpa_rasters/poly_marine1.shp')
  if(!file.exists(poly_marine_file)) {
    dir_rgn   <- file.path(dir_N, 'git-annex/globalprep/spatial/v2015/data')
    lyr_rgn <- 'regions_gcs'
    poly_rgn <- readShapePoly(file.path(dir_rgn, lyr_rgn), 
                               proj4string = crs_wgs84)

    ### filter out all polygons that are land or landlocked, and flag 
    ### the rest as spatially-marine = 1
    poly_marine <- poly_rgn[!str_detect(poly_rgn@data$rgn_typ, 'land'), ]
    poly_marine@data <- poly_marine@data %>%
      mutate(sp_mar = 1)

    poly_list <- list()
    poly_rows <- nrow(poly_marine)
    n_entities <- 20
    for (i in 1:ceiling(poly_rows/n_entities)) { # i = 1
      min_row <- (i - 1) * n_entities + 1
      max_row <- min(i * n_entities, poly_rows)
      poly_tmp <- poly_marine[min_row:max_row, ]
      message(sprintf('%s. Entities [%s:%s]', i, min_row, max_row))
      print(system.time({
        poly_tmp2 <- rgeos::gUnaryUnion(poly_tmp)
      })[3])
      poly_list[i] <- poly_tmp2
    }

    poly_mar_frankenstein <- spRbind(poly_list)

    # poly_mar_union <- rgeos::gUnaryUnion(poly_marine)

    # writePolyShape(poly_marine, poly_marine_file)
    writeOGR(obj    = poly_mar_frankenstein,
             dsn    = path.expand(dirname(poly_marine_file)),
             layer  = str_replace(basename(poly_marine_file), '.shp', ''),
             driver = 'ESRI Shapefile',
             overwrite_layer = TRUE,
             verbose = TRUE)

    } else {
    poly_marine <- readShapePoly(poly_marine_file, 
                               proj4string = crs_wgs84)
    }

  plot(poly_marine, col = 'blue')

  # rast_marine <- rasterize(poly_marine, rast_base, value = sp_mar_flag) 
  ### excruciatingly slow (interminable?)... consider gdal_rasterize?

  library(gdalUtils)
  rast_marine <- gdal_rasterize(
      src_datasource = poly_marine_file,
      dst_filename   = rast_marine_file, 
        # destination for output
      a = 'rgn_id', 
#      a = 'sp_mar', 
        # the attribute in the shapefile to be assigned to the cell values
      te = c(-180, -90, 180, 90), 
        # extents for output raster
      tr = c(.01, .01),  
        # resolution for output raster
      tap = TRUE, 
        # target aligned pixels - align coords of extent of output to values of -tr
      a_nodata = NA, 
        # nodata value for raster; otherwise they will be filled in with zeroes
      output_Raster = TRUE, 
        # return output as a RasterBrick? 
      verbose = TRUE)
}

@@

``` {r set_up_wdpa_raster}

### import WDPA database polygons from original .gdb
# gdb_wdpa   <- file.path(dir_N, 'git-annex/globalprep/_raw_data/wdpa_mpa/WDPA_Jan2015_Public/WDPA_Jan2015_Public.gdb')
# lyr_wdpa <- 'WDPA_poly_Jan2015'
# 
# poly_wdpa <- readOGR(dsn = gdb_wdpa, layer = lyr_wdpa, stringsAsFactors = FALSE)

### import WDPA database polygons from saved .shp
dir_wdpa   <- file.path(dir_N, 'git-annex/globalprep/_raw_data/wdpa_mpa/WDPA_Jan2015_Public')
lyr_wdpa <- 'WDPA_poly_Jan2015'
poly_wdpa <- readShapePoly(file.path(dir_wdpa, lyr_wdpa), 
                           proj4string = crs_wgs84)

prot_code_df <- data.frame('IUCN_CLASS' = c('I', 'IIa', 'IIb', 'III', 'IV', 'V', 'V', NA),
                           'prot_code'  = c( 1,    2,     2,     3,     4,   5,   6,   7))
poly_wdpa@data <- poly_wdpa@data %>%
  left_join(prot_code_df, by = 'IUCN_CLASS')

poly_mpa <- poly_wdpa[poly_wdpa@data$MARINE == 1, ]
poly_mpa <- poly_wdpa[poly_wdpa@data$prot_code <= 4, ]

>>>>

library(rgdal)
library(maptools)
library(raster)
library(readr)
library(tidyr)
library(dplyr)
library(stringr)
library(rgeos)

@@

  rast_empty <- raster(ext = extent(rast_loiczid), resolution = 0.02)
  rast_base  <- raster::area(rast_empty) ### takes a while...

@@

### set up some objects for WGS 84 projection
p4s_wgs84 <- '+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'
crs_wgs84 <- CRS(p4s_wgs84)

poly_dir <- file.path(dir_anx, 'explore_am_v_iucn/mpa_rasters')
rast_dir <- file.path(dir_anx, 'explore_am_v_iucn/mpa_rasters')

dir_rgn   <- file.path(dir_N, 'git-annex/globalprep/spatial/v2015/data')
lyr_rgn <- 'regions_gcs'
poly_marine <- readShapePoly(file.path(dir_rgn, lyr_rgn), 
                             proj4string = crs_wgs84)

### filter out all polygons that are land or landlocked, and flag 
### the rest as spatially-marine = 1

poly_marine <- poly_marine[!str_detect(poly_marine@data$rgn_typ, 'land'), ]
poly_list <- list()
poly_list[['poly_fao']]     <- poly_marine[str_detect(poly_marine@data$ant_typ, 'fao'), ]
poly_list[['poly_ccamlr']]  <- poly_marine[str_detect(poly_marine@data$ant_typ, 'ccamlr'), ]
poly_list[['poly_disp']]    <- poly_marine[str_detect(poly_marine@data$ant_typ, 'disputed'), ]
poly_eez    <- poly_marine[!(str_detect(poly_marine@data$ant_typ, 'ccamlr')) & 
                           !(str_detect(poly_marine@data$ant_typ, 'fao'))   &
                           !(str_detect(poly_marine@data$ant_typ, 'disputed')), ]
n_eez <- 5
for (i in 1:ceiling(nrow(poly_eez)/n_eez)) {
  min_row <- n_eez * (i - 1) + 1
  max_row <- n_eez * i
  poly_list[[paste('poly_eez', i, sep = '')]] <- poly_eez[min_row:max_row, ]
}

for (fn in names(poly_list)) { # fn = names(poly_list)[1] 
  message('Simplifying polygon ', fn)
  poly_list[[fn]] <- poly_simplify(poly_list[[fn]], thresh = 1e-5, verbose = TRUE)

#   if(!gIsValid(poly_list[[fn]], byid = TRUE, reason = TRUE)) {
#     message('INVALID POLYGONS FOUND in polygon ', fn)
#   }
  message('Writing polygon ', fn)
  writeOGR(obj    = poly_list[[fn]],
           dsn    = path.expand(poly_dir),
           layer  = fn,
           driver = 'ESRI Shapefile',
           overwrite_layer = TRUE,
           verbose = TRUE)
}

@@

``` {r turn saved files into rasters}
poly_names <- names(poly_list)
remove(poly_list, poly_eez, poly_marine)

ext  <- c(-180, -90, 180, 90), 
          # extents for output raster
reso <- c(.05, .05),

# tmpdir <- '~/home_big/ohara/R_raster_tmp'
# dir.create(tmpdir, showWarnings = FALSE)
# rasterOptions(tmpdir = tmpdir)

### gdalUtils::gdal_rasterize() method:
for (fn in poly_names) { # fn = poly_names[1]
  message('Rasterizing polygon ', fn)

  src <- path.expand(file.path(poly_dir, paste(fn, '.shp', sep = '')))
  rfn <- str_replace(fn, 'poly', 'rast') 
  tmp_dst <- file.path(rast_dir, 'tmp', paste(rfn, '.tif', sep = '')) %>%
    path.expand
  fnl_dst <- file.path(rast_dir, paste(rfn, '.tif', sep = '')) %>%
    path.expand

# if(!file.exists(fnl_dst)) {
#     if(file.size(src) > 7e6) {
#       message(sprintf('%s: Too damn big... %.2f', src, file.size(src)/1e6))
#     } else {
      ptm <- proc.time()
      tmp_rast <- gdalUtils::gdal_rasterize(
        src_datasource = src,
        dst_filename   = tmp_dst, 
          # destination for output
        a = 'rgn_id', 
          # the attribute in the shapefile to be assigned to the cell values
        te = ext, 
          # extents for output raster
        tr = reso,  
          # resolution for output raster
        a_nodata = NA, 
          # nodata value for raster; otherwise they will be filled in with zeroes
        output_Raster = TRUE,
          # return output as a RasterBrick? 
        verbose = TRUE)
      message('Rasterized ', rfn, '; ', (proc.time() - ptm)[3], ' seconds')

      ptm <- proc.time()
      raster::writeRaster(tmp_rast, fnl_dst, overwrite = TRUE)
      message('Write ', rfn, '; ', (proc.time() - ptm)[3], ' seconds')
      remove(tmp_rast)
      unlink(tmp_dst)
    # }
#   } else {
#     message('File ', fnl_dst, ' already exists.')
#   }
}

### 2016-02-25 mpa_analysis/mpa_analysis3.Rmd

    ### filter out all polygons that are land or landlocked, and flag 
    ### the rest as spatially-marine = 1
    poly_marine <- poly_rgn[!str_detect(poly_rgn@data$rgn_typ, 'land'), ]
    poly_marine@data <- poly_marine@data %>%
      mutate(sp_mar = 1)

    poly_list <- list()
    poly_rows <- nrow(poly_marine)
    n_entities <- 20
    for (i in 1:ceiling(poly_rows/n_entities)) { # i = 1
      min_row <- (i - 1) * n_entities + 1
      max_row <- min(i * n_entities, poly_rows)
      poly_tmp <- poly_marine[min_row:max_row, ]
      message(sprintf('%s. Entities [%s:%s]', i, min_row, max_row))
      print(system.time({
        poly_tmp2 <- rgeos::gUnaryUnion(poly_tmp)
      })[3])
      poly_list[i] <- poly_tmp2
    }

    poly_mar_frankenstein <- spRbind(poly_list)

    # poly_mar_union <- rgeos::gUnaryUnion(poly_marine)

    # writePolyShape(poly_marine, poly_marine_file)
    writeOGR(obj    = poly_mar_frankenstein,
             dsn    = path.expand(dirname(poly_marine_file)),
             layer  = str_replace(basename(poly_marine_file), '.shp', ''),
             driver = 'ESRI Shapefile',
             overwrite_layer = TRUE,
             verbose = TRUE)

    } else {
    poly_marine <- readShapePoly(poly_marine_file, 
                               proj4string = crs_wgs84)
    }

  plot(poly_marine, col = 'blue')

  # rast_marine <- rasterize(poly_marine, rast_base, value = sp_mar_flag) 
  ### excruciatingly slow (interminable?)... consider gdal_rasterize?

  library(gdalUtils)
  rast_marine <- gdal_rasterize(
      src_datasource = poly_marine_file,
      dst_filename   = rast_marine_file, 
        # destination for output
      a = 'rgn_id', 
#      a = 'sp_mar', 
        # the attribute in the shapefile to be assigned to the cell values
      te = c(-180, -90, 180, 90), 
        # extents for output raster
      tr = c(.01, .01),  
        # resolution for output raster
      tap = TRUE, 
        # target aligned pixels - align coords of extent of output to values of -tr
      a_nodata = NA, 
        # nodata value for raster; otherwise they will be filled in with zeroes
      output_Raster = TRUE, 
        # return output as a RasterBrick? 
      verbose = TRUE)
}

@@

prot_code_df <- data.frame('IUCN_CLASS' = c('I', 'IIa', 'IIb', 'III', 'IV', 'V', 'V', NA),
                           'prot_code'  = c( 1,    2,     2,     3,     4,   5,   6,   7))
poly_wdpa@data <- poly_wdpa@data %>%
  left_join(prot_code_df, by = 'IUCN_CLASS')

poly_mpa <- poly_wdpa[poly_wdpa@data$MARINE == 1, ]
poly_mpa <- poly_wdpa[poly_wdpa@data$prot_code <= 4, ]

>>>>

    ### filter out all polygons that are land or landlocked, and flag 
    ### the rest as spatially-marine = 1
    poly_marine <- poly_rgn[!str_detect(poly_rgn@data$rgn_typ, 'land'), ]

    # writePolyShape(poly_marine, poly_marine_file)
    writeOGR(obj    = poly_marine,
             dsn    = path.expand(dirname(poly_marine_file)),
             layer  = str_replace(basename(poly_marine_file), '.shp', ''),
             driver = 'ESRI Shapefile',
             overwrite_layer = TRUE,
             verbose = TRUE)
  }

  library(gdalUtils)
  rast_marine <- gdal_rasterize(
      src_datasource = poly_marine_file,
      dst_filename   = rast_marine_file, 
        # destination for output
      a = 'rgn_id', 
        # the attribute in the shapefile to be assigned to the cell values
      te = c(-180, -90, 180, 90), 
        # extents for output raster
      tr = c(.01, .01),  
        # resolution for output raster
      tap = TRUE, 
        # target aligned pixels - align coords of extent of output to values of -tr
      a_nodata = NA, 
        # nodata value for raster; otherwise they will be filled in with zeroes
      output_Raster = TRUE, 
        # return output as a RasterBrick? 
      verbose = TRUE)
}

@@

prot_code_df <- data.frame('IUCN_CLASS' = c('I', 'IIa', 'IIb', 'III', 'IV', 'V', 'V', NA),
                           'prot_code'  = c( 1,    2,     2,     3,     4,   5,   6,   7))
poly_wdpa@data <- poly_wdpa@data %>%
  left_join(prot_code_df, by = 'IUCN_CLASS') %>%
  mutate(prot_code = prot_code + MARINE)

### 2016-02-25 mpa_analysis/mpa_gap_analysis.Rmd

``` {r set_up_base_raster}
### DO WE NEED THIS? Use area at 0.5 degree level; most cells this will be pretty close.
### base raster will contain 0.01 degree cells from -180 to +180,
### and -90 to +90, with the associated area of each cell.
### Save it as a .tif in GitHub - about 50 
rast_base_file <- file.path(dir_git, 'mpa_analysis/rast_base.tif')

if(!file.exists(rast_base_file)) {
  ### import AquaMaps half-degree loiczid raster - for extents and such
  rast_loiczid_file   <- file.path(dir_git, 'data/rasters/loiczid_raster.grd')
  rast_loiczid        <- raster(rast_loiczid_file)
  names(rast_loiczid) <- 'loiczid'

  rast_empty <- raster(ext = extent(rast_loiczid), resolution = 0.01)
  rast_base  <- raster::area(rast_empty) ### takes a while...

  writeRaster(rast_base, rast_base_file, overwrite = TRUE)
} else {
  rast_base <- raster(rast_base_file)
}

@@

dir_rast <- file.path(dir_anx, 'explore_am_v_iucn/mpa_rasters')
rast_mpa_mar_file <- file.path(dir_rast, 'rast_mpa_mar.tif')
if(!file.exists(rast_mpa_mar_file)) {
  rast_mar_file <- file.path(dir_rast, 'rast_marine.tif')
  rast_mpa_file <- file.path(dir_rast, 'WDPA_poly_Jan2015_mpa1.tif')

@@

  rast_mpa_mar <- mask(rast_mpa, rast_mar)
  writeRaster(rast_mpa_mar, rast_mpa_mar_file)
} else {
  rast_mar            <- raster(rast_mar_file)
  names(rast_mar)     <- 'rgn_id'
  rast_mpa_mar        <- raster(rast_mpa_mar_file)
  names(rast_mpa_mar) <- 'status_yr'
}

@@

rast_loiczid_01_file <- file.path(dir_rast, 'rast_loiczid_01.tif')
rast_loiczid_01 <- resample(rast_loiczid, rast_mpa_mar, method = 'ngb')
writeRaster(rast_loiczid_01, rast_loiczid_01_file)

### Turn all marine cells to value of 1 (so can uze zonal sum)
rast_is_mar <- rast_mar
values(rast_is_mar)[!is.na(values(rast_mar))] <- 1
names(rast_is_mar) <- 'is_marine'

### Turn all MPA cells to value of 1 (so can use zonal sum)
### NOTE: loses status_yr values - so filter by year first
rast_is_mpa <- rast_mpa_mar
values(rast_is_mpa)[!is.na(values(rast_mpa_mar))] <- 1
names(rast_is_mpa) <- 'is_mpa'

rast_test <- rast_test2 <- raster(ext = extent(rast_loiczid), resolution = c(10, 10))
values(rast_test) <- 1:648
values(rast_test2) <- rep(c(1:10), each = 64, length.out = length(rast_test))
x <- zonal(rast_test, rast_test2, fun = sum)

>>>>

``` {r set_up_base_raster}
### DO WE NEED THIS? Use area at 0.5 degree level; most cells this will be pretty close.
### base raster will contain 0.01 degree cells from -180 to +180,
### and -90 to +90, with the associated area of each cell.
### Save it as a .tif in GitHub - about 50 
# rast_base_file <- file.path(dir_git, 'mpa_analysis/rast_base.tif')
# 
# if(!file.exists(rast_base_file)) {
#   ### import AquaMaps half-degree loiczid raster - for extents and such
#   rast_loiczid_file   <- file.path(dir_git, 'data/rasters/loiczid_raster.grd')
#   rast_loiczid        <- raster(rast_loiczid_file)
#   names(rast_loiczid) <- 'loiczid'
# 
#   rast_empty <- raster(ext = extent(rast_loiczid), resolution = 0.01)
#   rast_base  <- raster::area(rast_empty) ### takes a while...
#   
#   writeRaster(rast_base, rast_base_file,
#               progress = 'text', overwrite = TRUE)
# } else {
#   rast_base <- raster(rast_base_file)
# }

@@

dir_rast <- file.path(dir_anx, 'explore_am_v_iucn/mpa_rasters')
rast_mpa_mar_file <- file.path(dir_rast, 'rast_mpa_mar.tif')
rast_mar_file <- file.path(dir_rast, 'rast_marine.tif')

if(!file.exists(rast_mpa_mar_file)) {
  rast_mpa_file <- file.path(dir_rast, 'WDPA_poly_Jan2015_mpa1.tif')

@@

  rast_mpa_mar <- mask(rast_mpa, rast_mar, 
                       filename = rast_mpa_mar_file, 
                       overwrite = TRUE,
                       progress = 'text')
} else {
  rast_mar            <- raster(rast_mar_file)
  names(rast_mar)     <- 'rgn_id'
  rast_mpa_mar        <- raster(rast_mpa_mar_file)
  names(rast_mpa_mar) <- 'status_yr'
}

@@

rast_loiczid_01_file <- file.path(dir_rast, 'rast_loiczid_01.tif')
if(!file.exists(rast_loiczid_01_file)) {
  rast_loiczid_01 <- resample(rast_loiczid, rast_mpa_mar, 
                            method = 'ngb',
                            progress = 'text',
                            filename = rast_loiczid_01_file)
# writeRaster(rast_loiczid_01, rast_loiczid_01_file, progress = 'text')
} else {
  rast_loiczid_01 <- raster(rast_loiczid_01_file)
}
```

``` {r summarize counts of marine and MPAs by cell}

### Count marine cells in each LOICZID cell
mar_df_file <- file.path(dir_git, 'mpa_analysis/loiczid_marine.csv')
if(!file.exists(mar_df_file)) {
  is_marine_df <- zonal(rast_mar, 
                        z = rast_loiczid_01, 
                        fun = 'count',
                        progress = 'text') %>%
    as.data.frame() %>%
    rename(loiczid = zone)
  write_csv(is_marine_df, mar_df_file)
}

### Count all protected cells in each LOICZID cell
### NOTE: loses status_yr values - so filter by year for each iteration
yr_list <- c('y1990', 'y1995', 'y2000', 'y2005', 'y2010', 'y2014', 'y2015')
mpa_csvs_missing <- !file.exists(file.path(dir_git, sprintf('mpa_analysis/loiczid_mpa_%s.csv', yr_list)))

if(any(mpa_csvs_missing)) {
  yr_list1 <- yr_list[mpa_csvs_missing]
  message('Creating marine protected area rasters for years:')
  message('  ', paste(yr_list, collapse = ', '))
  for (yr in yr_list1) { # yr <- yr_list1[1]
    message('filtering MPAs to those designated prior to ', str_replace(yr, 'y', ''))
    rast_is_mpa <- rast_mpa_mar
    data_year <- as.integer(str_replace(yr, 'y', ''))
    values(rast_is_mpa)[values(rast_mpa_mar) >= data_year] <- NA
      ### NOTE: 2015 MPA dataset includes up to 2014; 2014 dataset goes
      ###   through 2013; etc.
    message('counting protected cells per LOICZID for ', yr)
    is_mpa_df <- zonal(rast_is_mpa, 
                       z = rast_loiczid_01, 
                       fun = 'count',
                       progress = 'text') %>%
      as.data.frame() %>%
      rename(loiczid = zone)
    write_csv(is_mpa_df, 
              file.path(dir_git, sprintf('mpa_analysis/loiczid_mpa_%s.csv', yr)))
  }
}
```

First step:

* Create data frame of LOICZID vs MPA coverage and marine area
    * note: using raster-generated marine area, not AquaMaps reported marine
      area - need to look into how different these are.
        * 7% of cells differ in calculated vs reported area
        * most of these differences cluster very near zero

For each species:

* for each "present" LOICZID cell
    * determine total area: overall cell area (in km2) * marine area (cells/2500)
    * determine protected area: overall cell area (in km2) * protected area (cells/2500)
* sum total area and protected area across all "present" LOICZID cells
* report: 
    * ratio of (protected area:total area) (dimensionless, percentage)
    * total marine area and total protected area (km2)
    * species name, ID, taxonomic group, extinction risk

``` {r combine_mpa_areas}

dir_am_data <- file.path(dir_N, 'git-annex/globalprep/_raw_data/aquamaps/v2015/csv/')

hcaf <- read_csv(file.path(dir_am_data, 'hcaf_truncated.csv')) %>%
  select(loiczid, cellarea, oceanarea)

cell_df <- read_csv(file.path(dir_git, 'mpa_analysis/loiczid_marine.csv'),
  col_types = 'dd') %>%
  rename(marine_ct = count) %>%
  left_join(hcaf, by = 'loiczid') %>%
  mutate(marine_pct  = marine_ct/2500,
         marine_area = marine_pct * cellarea) %>%
  select(-marine_ct)

for (yr in yr_list) { # yr <- yr_list[1]
  fn <- file.path(dir_git, sprintf('mpa_analysis/loiczid_mpa_%s.csv', yr))
  df <- read_csv(fn, col_types = 'dd')
  names(df)[2] <- yr
  cell_df <- cell_df %>%
    left_join(df, by = 'loiczid')
  cell_df[ , paste(yr, '_pct', sep = '')]  <- cell_df[ , yr]/2500
  cell_df[ , paste(yr, '_area', sep = '')] <- cell_df[ , paste(yr, '_pct', sep = '')] * cell_df$cellarea

  ### remove intermediate columns
  cell_df[ , yr] <- NULL
  cell_df[ , paste(yr, '_pct', sep = '')] <- NULL
}

# ### check diffs in oceanarea vs marine_area
# x <- cell_df %>% 
#   filter(oceanarea != marine_area) %>%
#   select(loiczid, oceanarea, marine_area, cellarea)
# ### 18158 cells out of 259200 = 7%
# x <- x %>% mutate(pct_diff = (marine_area - oceanarea)/cellarea)
# hist(x$pct_diff)
# ### most differences cluster right around zero.  Not going to worry about it!

``` {r apply loiczid area df to species list}

### Assemble the AM species-cell df with info on marine area and 
### protected area per cell
am_spp_cells <- read_csv(file.path(dir_am_data, 'hcaf_sp_native_trunc.csv'))
am_spp_cells <- am_spp_cells %>%
  left_join(cell_df, by = 'loiczid')

### filter to the appropriate threshold...
thresh <- 0.50

am_spp_cells_filt <- am_spp_cells %>%
  filter(probability >= thresh)

### group by speciesid, then summarize sum(marine area), sum(protected area), ratio
am_spp_mpas <- am_spp_cells %>%
  group_by(speciesid) %>%
  summarize(cellarea_tot = sum(cellarea), 
            ocean_range  = sum(oceanarea), 
            marine_range = sum(marine_area),
            mpa1990 = sum(y1990_area),
            mpa1995 = sum(y1995_area),
            mpa2000 = sum(y2000_area),
            mpa2005 = sum(y2005_area),
            mpa2010 = sum(y2010_area),
            mpa2014 = sum(y2014_area),
            mpa2015 = sum(y2015_area),
            pct1990 = mpa1990/marine_range,
            pct1995 = mpa1995/marine_range,
            pct2000 = mpa2000/marine_range,
            pct2005 = mpa2005/marine_range,
            pct2010 = mpa2010/marine_range,
            pct2014 = mpa2014/marine_range,
            pct2015 = mpa2015/marine_range)
```

``` {r recreate Klein results}

area_mpa <- cell_df %>%
  summarize(marine_area  = sum(marine_area), 
            mpa_tot_1990 = sum(y1990_area), 
            mpa_tot_1995 = sum(y1995_area), 
            mpa_tot_2000 = sum(y2000_area), 
            mpa_tot_2005 = sum(y2005_area), 
            mpa_tot_2010 = sum(y2010_area), 
            mpa_tot_2014 = sum(y2014_area), 
            mpa_tot_2015 = sum(y2015_area))
#   marine_area  mpa_tot_1990  mpa_tot_1995  mpa_tot_2000  mpa_tot_2005   mpa_tot_2010   mpa_tot_2014
#   362,657,307       351,592       532,098       748,172       932,817      1,137,142      2,885,465

### Recreate Klein results, using 2014 AquaMaps
mpa_klein_am2014 <- read_csv(file.path(dir_am_data, 'hcaf_sp_native_trunc_2014.csv'))
mpa_klein_am2014 <- mpa_klein_am2014 %>%
  left_join(cell_df, by = 'loiczid')

### filter to the appropriate threshold...
thresh <- 0.50

mpa_klein_am2014_filt <- mpa_klein_am2014 %>%
  filter(probability >= thresh)

### group by speciesid, then summarize sum(marine area), sum(protected area), ratio
mpa_klein_am2014 <- mpa_klein_am2014 %>%
  group_by(speciesid) %>%
  summarize(marine_range = sum(marine_area),
            mpa2014 = sum(y2014_area),
            pct2014 = mpa2014/marine_range)

mpa_klein_am2014 <- mpa_klein_am2014 %>%
  mutate(cat2014 = ifelse(pct2014 == 0,                       'gap', NA),
         cat2014 = ifelse(pct2014 > 0    & pct2014 <= 0.02,  '0-2%', cat2014),
         cat2014 = ifelse(pct2014 > 0.02 & pct2014 <= 0.05,  '2-5%', cat2014),
         cat2014 = ifelse(pct2014 > 0.05 & pct2014 <= 0.10, '5-10%', cat2014),
         cat2014 = ifelse(pct2014 > 0.10,                    '>10%', cat2014))

mpa_klein_orig <- mpa_klein_am2014 %>%
  group_by(cat2014) %>%
  summarize(n_tot = nrow(mpa_klein_am2014), n_cat = n(), pct_cat = round(n_cat/n_tot * 100, 1))

#   cat2014 n_tot n_cat pct_cat
#       gap 17348   378     2.2
#      0-2% 17348  8859    51.2
#      2-5% 17348  6728    38.7
#     5-10% 17348  1096     6.3
#      >10% 17348   287     1.7

### Redo analysis using 50% threshold, but new AquaMaps database
### * select just 2014 and 2015, at 50% threshold, and group by
###   same categories that the original paper did.
mpa_klein_am2015 <- am_spp_mpas %>%
  select(speciesid, marine_range, mpa2014, mpa2015, pct2014, pct2015)

mpa_klein_am2015 <- mpa_klein_am2015 %>%
  mutate(cat2014 = ifelse(pct2014 == 0,                       'gap', NA),
         cat2014 = ifelse(pct2014 > 0    & pct2014 <= 0.02,  '0-2%', cat2014),
         cat2014 = ifelse(pct2014 > 0.02 & pct2014 <= 0.05,  '2-5%', cat2014),
         cat2014 = ifelse(pct2014 > 0.05 & pct2014 <= 0.10, '5-10%', cat2014),
         cat2014 = ifelse(pct2014 > 0.10,                    '>10%', cat2014),
         cat2015 = ifelse(pct2015 == 0,                       'gap', NA),
         cat2015 = ifelse(pct2015 > 0    & pct2015 <= 0.02,  '0-2%', cat2015),
         cat2015 = ifelse(pct2015 > 0.02 & pct2015 <= 0.05,  '2-5%', cat2015),
         cat2015 = ifelse(pct2015 > 0.05 & pct2015 <= 0.10, '5-10%', cat2015),
         cat2015 = ifelse(pct2015 > 0.10,                    '>10%', cat2015))

mpa_klein_sum2014 <- mpa_klein_am2015 %>%
  group_by(cat2014) %>%
  summarize(n_tot = nrow(mpa_klein_am2015), n_cat = n(), pct_cat = round(n_cat/n_tot * 100, 1))

mpa_klein_sum2015 <- mpa_klein_am2015 %>%
  group_by(cat2015) %>%
  summarize(n_tot = nrow(mpa_klein_am2015), n_cat = n(), pct_cat = round(n_cat/n_tot * 100, 1))

#   cat2014 n_tot n_cat pct_cat
#       gap 22889   552    2.4
#      0-2% 22889 11546   50.4
#      2-5% 22889  8947   39.1
#     5-10% 22889  1456    6.4
#      >10% 22889   388    1.7

#   cat2015 n_tot n_cat pct_cat
#       gap 22889   552     2.4
#      0-2% 22889 11520    50.3
#      2-5% 22889  8960    39.1
#     5-10% 22889  1466     6.4
#      >10% 22889   391     1.7

### Original study (Klein 2015) vs new analyses

| study           | Protected areas    | Area (km^2^) | Gap  |  0–2% |  2–5% | 5–10% | >10% |
| :-------------: | :----------------- | :----------: | :--: | :---: | :---: | :---: | :--: |
| Klein 2015      | IUCN I-IV, Sp. Mar |   2,967,898  | 1.4% | 42.3% | 46.8% |  6.9% | 2.6% |
| am2014, mpa2014 | IUCN I-IV, Sp. Mar |   2,885,465  | 2.2% | 51.1% | 38.8% |  6.3% | 1.7% |
| am2015, mpa2015 | IUCN I-IV, Sp. Mar |   2,889,758  | 2.4% | 50.3% | 39.1% |  6.4% | 1.7% |
| am2015, mpa2014 | IUCN I-IV, Sp. Mar |   2,885,465  | 2.4% | 50.4% | 39.1% |  6.4% | 1.7% |
| am2015, mpa2000 | IUCN I-IV, Sp. Mar |     748,172  | 3.5% | 91.6% |  4.4% |  0.4% | 0.1% |
| am2015, mpa1990 | IUCN I-IV, Sp. Mar |     351,592  | 4.0% | 94.2% |  1.6% |  0.1% | 0.0% |

``` {r check results for 1990 and 2000}
mpa_spp <- am_spp_mpas %>%
  select(speciesid, marine_range, mpa1990, mpa2000, pct1990, pct2000)

mpa_spp <- mpa_spp %>%
  mutate(cat1990 = ifelse(pct1990 == 0,                       'gap', NA),
         cat1990 = ifelse(pct1990 > 0    & pct1990 <= 0.02,  '0-2%', cat1990),
         cat1990 = ifelse(pct1990 > 0.02 & pct1990 <= 0.05,  '2-5%', cat1990),
         cat1990 = ifelse(pct1990 > 0.05 & pct1990 <= 0.10, '5-10%', cat1990),
         cat1990 = ifelse(pct1990 > 0.10,                    '>10%', cat1990),
         cat1995 = ifelse(pct1995 == 0,                       'gap', NA),
         cat1995 = ifelse(pct1995 > 0    & pct1995 <= 0.02,  '0-2%', cat1995),
         cat1995 = ifelse(pct1995 > 0.02 & pct1995 <= 0.05,  '2-5%', cat1995),
         cat1995 = ifelse(pct1995 > 0.05 & pct1995 <= 0.10, '5-10%', cat1995),
         cat1995 = ifelse(pct1995 > 0.10,                    '>10%', cat1995),
         cat2000 = ifelse(pct2000 == 0,                       'gap', NA),
         cat2000 = ifelse(pct2000 > 0    & pct2000 <= 0.02,  '0-2%', cat2000),
         cat2000 = ifelse(pct2000 > 0.02 & pct2000 <= 0.05,  '2-5%', cat2000),
         cat2000 = ifelse(pct2000 > 0.05 & pct2000 <= 0.10, '5-10%', cat2000),
         cat2000 = ifelse(pct2000 > 0.10,                    '>10%', cat2000),
         cat2005 = ifelse(pct2005 == 0,                       'gap', NA),
         cat2005 = ifelse(pct2005 > 0    & pct2005 <= 0.02,  '0-2%', cat2005),
         cat2005 = ifelse(pct2005 > 0.02 & pct2005 <= 0.05,  '2-5%', cat2005),
         cat2005 = ifelse(pct2005 > 0.05 & pct2005 <= 0.10, '5-10%', cat2005),
         cat2005 = ifelse(pct2005 > 0.10,                    '>10%', cat2005),
         cat2010 = ifelse(pct2010 == 0,                       'gap', NA),
         cat2010 = ifelse(pct2010 > 0    & pct2010 <= 0.02,  '0-2%', cat2010),
         cat2010 = ifelse(pct2010 > 0.02 & pct2010 <= 0.05,  '2-5%', cat2010),
         cat2010 = ifelse(pct2010 > 0.05 & pct2010 <= 0.10, '5-10%', cat2010),
         cat2010 = ifelse(pct2010 > 0.10,                    '>10%', cat2010))

mpa_spp_1990 <- mpa_spp %>%
  group_by(cat1990) %>%
  summarize(n_tot = nrow(mpa_spp), n_cat = n(), pct_cat = round(n_cat/n_tot * 100, 1))

mpa_spp_2000 <- mpa_spp %>%
  group_by(cat2000) %>%
  summarize(n_tot = nrow(mpa_spp), n_cat = n(), pct_cat = round(n_cat/n_tot * 100, 1))

#   cat1990 n_tot n_cat pct_cat
#       gap 22889   925     4.0
#      0-2% 22889 21569    94.2
#      2-5% 22889   376     1.6
#     5-10% 22889    18     0.1
#      >10% 22889     1     0.0
# 
#   cat2000 n_tot n_cat pct_cat
#       gap 22889   801     3.5
#      0-2% 22889 20973    91.6
#      2-5% 22889  1004     4.4
#     5-10% 22889    92     0.4
#      >10% 22889    19     0.1
